---
title: "dplyr 1.2.0 Conversion Cheatsheet"
subtitle: "From Old Patterns to New Functions"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(palmerpenguins)
```

This cheatsheet walks through converting old dplyr patterns to the new dplyr 1.2.0 functions. Each section shows the old way and the new way side-by-side.

---

# 1. Dropping Rows: `filter()` → `filter_out()`

## Old Way

```{r}
# Using ! negation
penguins |>
  filter(!(species == "Adelie")) |>
  count(species)
```

## New Way

```{r}
# Clear intent: drop these rows
penguins |>
  filter_out(species == "Adelie") |>
  count(species)
```

## Why Better?

- ✅ Clear intent (drop vs keep)
- ✅ Better NA handling (keeps NAs by default)
- ✅ No complex negation logic needed

---

# 2. OR Logic: `|` → `when_any()`

## Old Way

```{r}
# Multiple | operators
penguins |>
  filter(
    (species == "Adelie" & island == "Torgersen" & body_mass_g > 3700) |
    (species == "Gentoo" & body_mass_g > 5000)
  ) |>
  select(species, island, body_mass_g)
```

## New Way

```{r}
# Clean OR logic with when_any()
penguins |>
  filter(
    when_any(
      when_all(species == "Adelie", island == "Torgersen", body_mass_g > 3700),
      when_all(species == "Gentoo", body_mass_g > 5000)
    )
  ) |>
  select(species, island, body_mass_g)
```

## Why Better?

- ✅ Explicit OR logic (no need to mentally parse `|`)
- ✅ Nests cleanly with `when_all()`
- ✅ Easier to add/remove conditions

---

# 3. Simple OR: `|` → `when_any()`

## Old Way

```{r}
# Simple OR with |
penguins |>
  filter(species == "Adelie" | body_mass_g > 5000) |>
  select(species, island, body_mass_g)
```

## New Way

```{r}
# Explicit OR
penguins |>
  filter(when_any(species == "Adelie", body_mass_g > 5000)) |>
  select(species, island, body_mass_g)
```

## Why Better?

- ✅ Clearer intent (especially for newcomers)
- ✅ Consistent with nested `when_any()` patterns

---

# 4. Value Recoding: `case_when()` → `recode_values()`

## Old Way

```{r}
# Repetitive == checks
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_when(
      island == "Biscoe" ~ "Southwest",
      island == "Dream" ~ "Northwest",
      island == "Torgersen" ~ "Northwest",
      .default = island
    )
  )
```

## New Way

```{r}
# Direct value mapping
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = island |>
      recode_values(
        "Biscoe" ~ "Southwest",
        "Dream" ~ "Northwest",
        "Torgersen" ~ "Northwest"
      )
  )
```

## Why Better?

- ✅ No repetitive `==` checks
- ✅ Clearer intent (recoding values)
- ✅ Can use lookup tables
- ✅ Better error handling with `unmatched = "error"`

---

# 5. Using Lookup Tables: `case_when()` → `recode_values()`

## Old Way

```{r}
# Manual case_when from vectors
island_mapping <- list(
  "Biscoe" = "Southwest",
  "Dream" = "Northwest",
  "Torgersen" = "Northwest"
)

# Have to construct case_when manually
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_when(
      island == "Biscoe" ~ island_mapping[["Biscoe"]],
      island == "Dream" ~ island_mapping[["Dream"]],
      island == "Torgersen" ~ island_mapping[["Torgersen"]],
      .default = island
    )
  )
```

## New Way

```{r}
# Use tibble directly as lookup
island_mapping_tibble <- tibble(
  from = c("Biscoe", "Dream", "Torgersen"),
  to = c("Southwest", "Northwest", "Northwest")
)

penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = island |>
      recode_values(
        from = island_mapping_tibble$from,
        to = island_mapping_tibble$to
      )
  )
```

## Why Better?

- ✅ Lookup tables stored as data frames
- ✅ Easy to maintain and update mappings
- ✅ Can load from CSV or database

---

# 6. Conditional Replacement: `case_when()` → `replace_when()`

## Old Way

```{r}
# Need .default for same column
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    body_mass_g = case_when(
      body_mass_g > 5000 ~ 5000,
      .default = body_mass_g  # Boilerplate!
    )
  )
```

## New Way

```{r}
# No .default needed!
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    body_mass_g = replace_when(
      body_mass_g,
      body_mass_g > 5000 ~ 5000
    )
  )
```

## Why Better?

- ✅ No `.default` boilerplate
- ✅ Clear intent (replacing specific values)
- ✅ Unmatched values automatically preserved

---

# 7. Simple Value Replacement: `case_when()` → `replace_values()`

## Old Way

```{r}
# Need .default even for exact match
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_when(
      island == "Biscoe" ~ "Southwest",
      island == "Dream" ~ "Northwest",
      island == "Torgersen" ~ "Northwest"
    )
  ) |>
  mutate(
    island_location = case_when(
      island_location == "Northwest" ~ "Northern Palmer Archipelago",
      .default = island_location
    )
  )
```

## New Way

```{r}
# Direct replacement
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = island |>
      recode_values(
        "Biscoe" ~ "Southwest",
        "Dream" ~ "Northwest",
        "Torgersen" ~ "Northwest"
      )
  ) |>
  mutate(
    island_location = island_location |>
      replace_values(
        "Northwest" ~ "Northern Palmer Archipelago"
      )
  )
```

## Why Better?

- ✅ No `.default` needed
- ✅ Clear that only specific values are being replaced
- ✅ Works with any number of replacements

---

# 8. Error on Unmatched: `case_when()` → `recode_values()`

## Old Way

```{r}
#| error: true
# Manual validation needed
islands <- c("Biscoe", "Dream", "UnknownIsland")
data <- tibble(island = islands)

# case_when() silently returns NA for unmatched
result <- data |>
  mutate(
    location = case_when(
      island == "Biscoe" ~ "Southwest",
      island == "Dream" ~ "Northwest"
      # UnknownIsland becomes NA - no error!
    )
  )

print(result)

# Would need to manually check for NAs
if (any(is.na(result$location))) {
  stop("Unmatched values found!")
}
```

## New Way

```{r}
#| error: true
# Automatic validation
islands <- c("Biscoe", "Dream", "UnknownIsland")
data <- tibble(island = islands)

# Errors immediately on unmatched value
data |>
  mutate(
    location = island |>
      recode_values(
        "Biscoe" ~ "Southwest",
        "Dream" ~ "Northwest",
        unmatched = "error"  # Explicit error handling
      )
  )
```

## Why Better?

- ✅ Fail fast on data issues
- ✅ Explicit error handling
- ✅ No silent NA creation

---

# Quick Reference Table

| Pattern | Old Way | New Way | Key Benefit |
|---------|---------|---------|-------------|
| Drop rows | `filter(!(x))` | `filter_out(x)` | Clear intent, better NA handling |
| OR logic | `a \| b` | `when_any(a, b)` | Explicit, nests cleanly |
| AND logic | `a & b` | `when_all(a, b)` | Explicit, nests with `when_any()` |
| Recode values | `case_when(x == "a" ~ "A", ...)` | `recode_values("a" ~ "A", ...)` | No repetitive `==` |
| Use lookup table | Manual `case_when()` | `recode_values(from = ..., to = ...)` | Direct data frame usage |
| Replace values conditionally | `case_when(..., .default = x)` | `replace_when(x, ...)` | No `.default` boilerplate |
| Replace specific values | `case_when(x == "a" ~ "A", .default = x)` | `replace_values("a" ~ "A")` | No `.default` boilerplate |
| Error on unmatched | Manual validation | `recode_values(..., unmatched = "error")` | Fail fast |

---

# Rules of Thumb

**Use `filter()` to keep rows**

**Use `filter_out()` to drop rows** (if you're using `!` in `filter()`)

**Use `when_any()` for OR logic** (if you have complex `|` expressions)

**Use `when_all()` for AND logic** (useful when nesting inside `when_any()`)

**Use `recode_values()` for exact matches** (if you're using `==` in `case_when()`)

**Use `replace_when()` to modify existing columns** (instead of `case_when()` with `.default`)

**Use `replace_values()` for simple replacements** (instead of `case_when()` with `.default`)

---

# Installation

```r
install.packages("pak")
pak::pak("dplyr")
```

Current version: dplyr 1.2.0

---

*Generated for dplyr 1.2.0 - Working smarter with new functions*
