---
title: "Exercises: Pre-dplyr 1.2.0 Patterns"
subtitle: "Rewrite these using dplyr 1.2.0's new functions — see solutions.qmd for answers"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(palmerpenguins)
```

Each exercise below uses a **pre-dplyr 1.2.0** approach. Your task is to rewrite it
using the corresponding new function from dplyr 1.2.0. Every exercise has a matching
solution in `solutions.qmd` (Exercise 1 → Solution 1, and so on).

---

## Data Overview

The exercises use the `penguins` dataset from the `palmerpenguins` package and
several small samples drawn from it. Here are some quick counts to help you reason
about expected results.

```{r}
#| label: data-overview
glimpse(penguins)
```

```{r}
#| label: counts-species
count(penguins, species)
```

```{r}
#| label: counts-island
count(penguins, island)
```

```{r}
#| label: counts-species-island
count(penguins, species, island)
```

```{r}
#| label: body-mass-range
penguins |>
  summarise(
    min_body_mass = min(body_mass_g, na.rm = TRUE),
    max_body_mass = max(body_mass_g, na.rm = TRUE),
    n_above_5000   = sum(body_mass_g > 5000, na.rm = TRUE),
    n_missing      = sum(is.na(body_mass_g))
  )
```

---

# Part 1: `filter()` Family

---

## Exercise 1: Dropping rows with complex conditions

This small sample mirrors the animation from the presentation. It includes an `NA`
in `body_mass_g` that makes negation-based filtering tricky.

```{r}
#| label: ex1-data
penguins_small <- tibble(
  species     = rep("Adelie", 5),
  island      = c("Torgersen", "Biscoe", "Dream", "Torgersen", "Torgersen"),
  body_mass_g = c(3750, 3800, 3300, 4500, NA_real_)
)

penguins_small
```

**Task:** Drop rows where **both** `island == "Torgersen"` **and**
`body_mass_g > 4000` are true. Rows where either condition is missing (`NA`) should
be *kept*.

Expected rows remaining after the operation:

| species | island    | body_mass_g | Why kept                                              |
|---------|-----------|-------------|-------------------------------------------------------|
| Adelie  | Biscoe    | 3800        | Island ≠ Torgersen                                    |
| Adelie  | Dream     | 3300        | Island ≠ Torgersen                                    |
| Adelie  | Torgersen | NA          | body_mass_g is NA — can't confirm it's > 4000, keep it |

```{r}
#| label: ex1-old
penguins_small |>
  filter(
    !((island == "Torgersen" & !is.na(island)) &
        (body_mass_g > 4000 & !is.na(body_mass_g)))
  )
```

> **Your turn:** Rewrite using `filter_out()` — see Solution 1.

---

## Exercise 2: Simple OR filtering

**Task:** From the full `penguins` dataset, keep rows where `species == "Adelie"` **or**
`body_mass_g > 5000`.

```{r}
#| label: ex2-context
# Helpful counts — how many rows do we expect?
cat("Adelie rows:", sum(penguins$species == "Adelie", na.rm = TRUE), "\n")
cat("body_mass_g > 5000 (non-Adelie):",
    sum(penguins$body_mass_g > 5000 & penguins$species != "Adelie", na.rm = TRUE), "\n")
```

```{r}
#| label: ex2-old
penguins |>
  filter(species == "Adelie" | body_mass_g > 5000) |>
  select(species, island, body_mass_g)
```

> **Your turn:** Rewrite using `filter()` + `when_any()` — see Solution 2.

---

## Exercise 3: Complex OR filtering with grouped conditions

This sample matches the presentation animation. Each group of conditions must all be
true together before the OR is applied.

```{r}
#| label: ex3-data
penguins_filters <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Adelie",
                  "Gentoo", "Gentoo", "Gentoo", "Gentoo"),
  island      = c("Torgersen", "Torgersen", "Dream", "Biscoe",
                  "Biscoe",    "Dream",     "Biscoe", "Dream"),
  body_mass_g = c(3750, 3800, 3250, NA_real_,
                  4500, 5700, 4450, 5700)
)

penguins_filters
```

**Task:** Keep rows where:
- **Group A:** `species == "Adelie"` AND `island == "Torgersen"` AND `body_mass_g > 3700`, **or**
- **Group B:** `species == "Gentoo"` AND `body_mass_g > 5000`

Expected result (4 rows):

| species | island    | body_mass_g | Matched group |
|---------|-----------|-------------|---------------|
| Adelie  | Torgersen | 3750        | A             |
| Adelie  | Torgersen | 3800        | A             |
| Gentoo  | Dream     | 5700        | B             |
| Gentoo  | Dream     | 5700        | B             |

```{r}
#| label: ex3-old
penguins_filters |>
  filter(
    (species == "Adelie" & island == "Torgersen" & body_mass_g > 3700) |
    (species == "Gentoo" & body_mass_g > 5000)
  )
```

> **Your turn:** Rewrite using `when_any()` nested inside `when_all()` — see Solution 3.

---

## Exercise 4: AND filtering

This sample mirrors the `when_all()` animation.

```{r}
#| label: ex4-data
penguins_when_all <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Adelie",
                  "Gentoo", "Gentoo", "Gentoo", "Gentoo"),
  island      = c("Torgersen", "Dream", "Biscoe", "Torgersen",
                  "Biscoe",    "Dream", "Biscoe", "Dream"),
  body_mass_g = c(3750, 3800, 3250, NA_real_,
                  4500, 5700, 4450, 5700)
)

penguins_when_all
```

**Task:** Keep rows where `species == "Gentoo"` **and** `body_mass_g > 5000`.

Expected result (2 rows):

| species | island | body_mass_g |
|---------|--------|-------------|
| Gentoo  | Dream  | 5700        |
| Gentoo  | Dream  | 5700        |

```{r}
#| label: ex4-old
penguins_when_all |>
  filter(species == "Gentoo" & body_mass_g > 5000)
```

> **Your turn:** Rewrite using `when_all()` — see Solution 4.

---

# Part 2: Recoding Values

---

## Exercise 5: Recode island names to locations

This 4-row sample matches the island-recoding animation.

```{r}
#| label: ex5-data
penguins_islands <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Gentoo"),
  island      = c("Biscoe", "Dream", "Torgersen", "Biscoe"),
  body_mass_g = c(3750, 3800, 4750, 5000)
)

penguins_islands
```

**Mapping:**

| island    | location  |
|-----------|-----------|
| Biscoe    | Southwest |
| Dream     | Northwest |
| Torgersen | Northwest |

**Task:** Add an `island_location` column using this mapping.

```{r}
#| label: ex5-old
penguins_islands |>
  mutate(
    island_location = case_when(
      island == "Biscoe"    ~ "Southwest",
      island == "Dream"     ~ "Northwest",
      island == "Torgersen" ~ "Northwest",
      .default = island
    )
  )
```

> **Your turn:** Rewrite using `recode_values()` inside `mutate()` — see Solution 5.

---

## Exercise 6: Recode using a lookup table

Same task as Exercise 5, but this time the mapping lives in a separate lookup table
instead of being written inline.

```{r}
#| label: ex6-data
# Same source data as Exercise 5
penguins_islands

# Lookup table
island_lookup <- tibble(
  from = c("Biscoe", "Dream", "Torgersen"),
  to   = c("Southwest", "Northwest", "Northwest")
)

island_lookup
```

**Task:** Add `island_location` using `island_lookup` as the mapping source.

```{r}
#| label: ex6-old
# Manual: pull each pair out of the lookup table one-by-one
penguins_islands |>
  mutate(
    island_location = case_when(
      island == island_lookup$from[island_lookup$from == "Biscoe"]    ~
        island_lookup$to[island_lookup$from == "Biscoe"],
      island == island_lookup$from[island_lookup$from == "Dream"]     ~
        island_lookup$to[island_lookup$from == "Dream"],
      island == island_lookup$from[island_lookup$from == "Torgersen"] ~
        island_lookup$to[island_lookup$from == "Torgersen"],
      .default = island
    )
  )
```

> **Your turn:** Rewrite using `recode_values()` with `from =` and `to =` — see Solution 6.

---

## Exercise 7: Replace specific values in an existing column

Starting from a dataset that already has an `island_location` column, replace one of
the location values with a longer name. This sample mirrors the `replace_values()`
animation.

```{r}
#| label: ex7-data
penguins_locations <- tibble(
  species        = c("Adelie", "Adelie", "Adelie", "Gentoo"),
  island         = c("Biscoe", "Dream", "Torgersen", "Biscoe"),
  body_mass_g    = c(3750, 3800, 4750, 5000),
  island_location = c("Southwest", "Northwest", "Northwest", "Southwest")
)

penguins_locations
```

**Task:** In `island_location`, replace `"Northwest"` with
`"Northern Palmer Archipelago"`. Leave `"Southwest"` unchanged.

Expected result:

| island_location             |
|-----------------------------|
| Southwest                   |
| Northern Palmer Archipelago |
| Northern Palmer Archipelago |
| Southwest                   |

```{r}
#| label: ex7-old
penguins_locations |>
  mutate(
    island_location = case_when(
      island_location == "Northwest" ~ "Northern Palmer Archipelago",
      .default = island_location   # boilerplate — must name the same column
    )
  )
```

> **Your turn:** Rewrite using `replace_values()` — see Solution 7.

---

## Exercise 8: Cap a numeric column at a threshold

This sample mirrors the `replace_when()` animation.

```{r}
#| label: ex8-data
penguins_large <- tibble(
  species     = c("Adelie", "Adelie", "Gentoo", "Gentoo"),
  island      = c("Biscoe", "Dream", "Biscoe", "Biscoe"),
  body_mass_g = c(3750, 3800, 5200, 5400)
)

penguins_large
```

**Task:** Cap `body_mass_g` at 5000 g — any value over 5000 should become 5000.
Values at or below 5000 stay unchanged.

```{r}
#| label: ex8-old
penguins_large |>
  mutate(
    body_mass_g = case_when(
      body_mass_g > 5000 ~ 5000,
      .default = body_mass_g   # boilerplate — must name the same column
    )
  )
```

> **Your turn:** Rewrite using `replace_when()` — see Solution 8.
