---
title: "Old Patterns - Pre-dplyr 1.2.0"
format: html
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(palmerpenguins)
```

This document shows working examples of the "old" ways of doing common data manipulation tasks that dplyr 1.2.0's new functions are designed to improve.

## Filtering Out Rows (Old Way)

**Problem:** Using `!` negation with `filter()` to drop rows

```{r}
# Drop rows where species is Adelie
penguins |>
  filter(!(species == "Adelie")) |>
  count(species)
```

**Problem with NA handling:**

```{r}
# Create data with NA
penguins_na <- penguins |>
  mutate(island = if_else(row_number() == 1, NA_character_, island))

# Using ! drops both the target AND NAs!
penguins_na |>
  filter(!(island == "Torgersen" & body_mass_g > 4000))
```

**Correct but complex workaround:**

```{r}
penguins_na |>
  filter(
    !((island == "Torgersen" & !is.na(island)) &
      (body_mass_g > 4000 & !is.na(body_mass_g)))
  )
```

## Complex OR Logic (Old Way)

**Problem:** Multiple `|` operators become hard to read

```{r}
# Keep Adelie from Torgersen with body mass > 3700g OR Gentoo with body mass > 5000g
penguins |>
  filter(
    (species == "Adelie" & island == "Torgersen" & body_mass_g > 3700) |
    (species == "Gentoo" & body_mass_g > 5000)
  ) |>
  select(species, island, body_mass_g)
```

## Complex AND Logic (Old Way)

**Problem:** Nested conditions are hard to track

```{r}
# Keep only Adelie from Torgersen with body mass > 3700g
penguins |>
  filter(
    species == "Adelie" &
    island == "Torgersen" &
    body_mass_g > 3700
  ) |>
  select(species, island, body_mass_g)
```

**Nested OR and AND:**

```{r}
# Complex: (Adelie from Torgersen with mass > 3700) OR (Gentoo from Biscoe)
penguins |>
  filter(
    ((species == "Adelie" & island == "Torgersen") & body_mass_g > 3700) |
    (species == "Gentoo" & island == "Biscoe")
  ) |>
  select(species, island, body_mass_g)
```

## Recoding Values with case_when() (Old Way)

**Problem:** Using `case_when()` with multiple `==` checks for simple value mapping

```{r}
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_when(
      island == "Biscoe" ~ "Southwest",
      island == "Dream" ~ "Northwest",
      island == "Torgersen" ~ "Northwest",
      .default = island
    )
  )
```

**Problem with unmatched values:**

```{r}
# Unmatched values become NA
penguins |>
  slice(1:5) |>
  select(species, island, body_mass_g) |>
  mutate(
    size = case_when(
      body_mass_g > 4500 ~ "Large",
      body_mass_g < 3500 ~ "Small"
      # Missing .default, so 3500-4500 becomes NA
    )
  )
```

## Using case_match() (Old Way - Now Soft Deprecated)

**Problem:** `case_match()` is being deprecated in favor of `recode_values()`

```{r}
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_match(
      island,
      "Biscoe" ~ "Southwest",
      "Dream" ~ "Northwest",
      "Torgersen" ~ "Northwest",
      .default = island
    )
  )
```

## Replacing Values with case_when() (Old Way)

**Problem:** Using `case_when()` with `.default` for simple replacements

```{r}
# Cap body_mass_g at 5000g
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    body_mass_g_capped = case_when(
      body_mass_g > 5000 ~ 5000,
      .default = body_mass_g  # Must specify .default even though it's the same column
    )
  )
```

**Even for single replacements:**

```{r}
# Replace just Northwest with a longer name
penguins |>
  slice(1:4) |>
  select(species, island, body_mass_g) |>
  mutate(
    island_location = case_when(
      island == "Biscoe" ~ "Southwest",
      island == "Dream" ~ "Northwest",
      island == "Torgersen" ~ "Northwest"
    )
  ) |>
  mutate(
    island_location = case_when(
      island_location == "Northwest" ~ "Northern Palmer Archipelago",
      .default = island_location  # Boilerplate!
    )
  )
```

## Summary of Pain Points

1. **`filter()` with `!`**: Complex NA handling, unclear intent
2. **Complex `|` and `&`**: Hard to read nested logic
3. **`case_when()` for exact matches**: Repetitive `==` checks
4. **`case_when()` for replacements**: Unnecessary `.default` boilerplate
5. **`case_match()`**: Being deprecated, used for simple value mapping

All of these patterns work, but dplyr 1.2.0 provides clearer, more expressive alternatives!
