---
title: "Solutions: dplyr 1.2.0 New Functions"
subtitle: "Solutions to exercises.qmd — numbered to match"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
---

```{r}
#| label: setup
#| message: false
library(dplyr)
library(palmerpenguins)
```

Each solution below corresponds directly to the exercise with the same number in
`exercises.qmd`. The task description is repeated for reference.

---

## Data Overview

```{r}
#| label: data-overview
glimpse(penguins)
```

```{r}
#| label: counts-species
count(penguins, species)
```

```{r}
#| label: counts-island
count(penguins, island)
```

```{r}
#| label: counts-species-island
count(penguins, species, island)
```

```{r}
#| label: body-mass-range
penguins |>
  summarise(
    min_body_mass = min(body_mass_g, na.rm = TRUE),
    max_body_mass = max(body_mass_g, na.rm = TRUE),
    n_above_5000  = sum(body_mass_g > 5000, na.rm = TRUE),
    n_missing     = sum(is.na(body_mass_g))
  )
```

---

# Part 1: `filter()` Family

---

## Solution 1 → Exercise 1: `filter_out()`

**Task:** Drop rows where `island == "Torgersen"` **and** `body_mass_g > 4000` are
both true. Rows with `NA` should be kept.

```{r}
#| label: sol1-data
penguins_small <- tibble(
  species     = rep("Adelie", 5),
  island      = c("Torgersen", "Biscoe", "Dream", "Torgersen", "Torgersen"),
  body_mass_g = c(3750, 3800, 3300, 4500, NA_real_)
)
```

```{r}
#| label: sol1-new
# filter_out() drops rows where ALL listed conditions are TRUE.
# Rows where any condition is NA are kept — no extra is.na() calls needed.
penguins_small |>
  filter_out(island == "Torgersen", body_mass_g > 4000)
```

**Why `filter_out()` is better:**

- Intent is explicit: you're *dropping*, not keeping
- `NA` rows are preserved automatically
- No need to negate (`!`) or manually guard against `NA`s

---

## Solution 2 → Exercise 2: `when_any()` (simple OR)

**Task:** Keep rows where `species == "Adelie"` **or** `body_mass_g > 5000`.

```{r}
#| label: sol2-new
penguins |>
  filter(when_any(
    species == "Adelie",
    body_mass_g > 5000
  )) |>
  select(species, island, body_mass_g)
```

**Why `when_any()` is better:**

- Makes the OR logic explicit and readable at a glance
- Each condition sits on its own line — easy to add or remove conditions
- Consistent syntax with the nested `when_any(when_all(...))` pattern (Solution 3)

---

## Solution 3 → Exercise 3: `when_any()` + `when_all()` (complex OR)

**Task:** Keep rows where:
- **Group A:** `species == "Adelie"` AND `island == "Torgersen"` AND `body_mass_g > 3700`, **or**
- **Group B:** `species == "Gentoo"` AND `body_mass_g > 5000`

```{r}
#| label: sol3-data
penguins_filters <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Adelie",
                  "Gentoo", "Gentoo", "Gentoo", "Gentoo"),
  island      = c("Torgersen", "Torgersen", "Dream", "Biscoe",
                  "Biscoe",    "Dream",     "Biscoe", "Dream"),
  body_mass_g = c(3750, 3800, 3250, NA_real_,
                  4500, 5700, 4450, 5700)
)
```

```{r}
#| label: sol3-new
# when_all() bundles AND conditions into one unit.
# when_any() then applies OR across those bundles.
penguins_filters |>
  filter(
    when_any(
      when_all(species == "Adelie", island == "Torgersen", body_mass_g > 3700),
      when_all(species == "Gentoo", body_mass_g > 5000)
    )
  )
```

**Why `when_any()` + `when_all()` is better:**

- Groups AND conditions together visually instead of relying on `()` nesting
- The two groups are clearly separated — easy to add a third group
- Matches how you'd describe the logic in plain English: "any of these groups"

---

## Solution 4 → Exercise 4: `when_all()` (AND)

**Task:** Keep rows where `species == "Gentoo"` **and** `body_mass_g > 5000`.

```{r}
#| label: sol4-data
penguins_when_all <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Adelie",
                  "Gentoo", "Gentoo", "Gentoo", "Gentoo"),
  island      = c("Torgersen", "Dream", "Biscoe", "Torgersen",
                  "Biscoe",    "Dream", "Biscoe", "Dream"),
  body_mass_g = c(3750, 3800, 3250, NA_real_,
                  4500, 5700, 4450, 5700)
)
```

```{r}
#| label: sol4-new
# when_all() makes AND logic explicit.
penguins_when_all |>
  filter(when_all(species == "Gentoo", body_mass_g > 5000))
```

**Why `when_all()` is better:**

- Symmetrical with `when_any()` — one consistent set of verbs for boolean logic
- Easy to nest inside `when_any()` when your logic grows more complex
- For simple AND, commas in `filter()` still work fine too

> **Rule of thumb:** For simple `&` conditions, plain commas in `filter()` are
> fine. Reach for `when_all()` when you need to nest it inside `when_any()`.

---

# Part 2: Recoding Values

---

## Solution 5 → Exercise 5: `recode_values()` inline

**Task:** Add an `island_location` column using the mapping
Biscoe → Southwest, Dream → Northwest, Torgersen → Northwest.

```{r}
#| label: sol5-data
penguins_islands <- tibble(
  species     = c("Adelie", "Adelie", "Adelie", "Gentoo"),
  island      = c("Biscoe", "Dream", "Torgersen", "Biscoe"),
  body_mass_g = c(3750, 3800, 4750, 5000)
)
```

```{r}
#| label: sol5-new
# recode_values() uses formula syntax: "old_value" ~ "new_value"
# No repetitive == checks needed.
penguins_islands |>
  mutate(
    island_location = island |>
      recode_values(
        "Biscoe"    ~ "Southwest",
        "Dream"     ~ "Northwest",
        "Torgersen" ~ "Northwest"
      )
  )
```

**Why `recode_values()` is better:**

- No repeated `island ==` on every line
- Formula syntax (`"old" ~ "new"`) clearly states input → output
- Unmatched values are `NA` by default; use `unmatched = "error"` to catch surprises

---

## Solution 6 → Exercise 6: `recode_values()` with a lookup table

**Task:** Same island → location mapping, but sourced from the `island_lookup` table.

```{r}
#| label: sol6-data
island_lookup <- tibble(
  from = c("Biscoe", "Dream", "Torgersen"),
  to   = c("Southwest", "Northwest", "Northwest")
)

island_lookup
```

```{r}
#| label: sol6-new
# Pass the lookup table's columns directly to from = / to =
penguins_islands |>
  mutate(
    island_location = island |>
      recode_values(
        from = island_lookup$from,
        to   = island_lookup$to
      )
  )
```

**Why `recode_values()` with a lookup table is better:**

- The mapping lives in a data frame — easy to load from a CSV or database
- Adding or changing a mapping means editing the table, not the pipeline
- Works seamlessly with `from =` / `to =` vectors of any length

---

## Solution 7 → Exercise 7: `replace_values()`

**Task:** In `island_location`, replace `"Northwest"` with
`"Northern Palmer Archipelago"`. Leave other values unchanged.

```{r}
#| label: sol7-data
penguins_locations <- tibble(
  species         = c("Adelie", "Adelie", "Adelie", "Gentoo"),
  island          = c("Biscoe", "Dream", "Torgersen", "Biscoe"),
  body_mass_g     = c(3750, 3800, 4750, 5000),
  island_location = c("Southwest", "Northwest", "Northwest", "Southwest")
)
```

```{r}
#| label: sol7-new
# replace_values() replaces exact matches in place.
# Unmatched values are left as-is — no .default needed.
penguins_locations |>
  mutate(
    island_location = island_location |>
      replace_values("Northwest" ~ "Northern Palmer Archipelago")
  )
```

**Why `replace_values()` is better:**

- No `.default = island_location` boilerplate
- Intent is clear: you're *replacing* a value, not rebuilding the column
- Use multiple formulas to replace several values in one call

---

## Solution 8 → Exercise 8: `replace_when()`

**Task:** Cap `body_mass_g` at 5000 g. Values above 5000 become 5000;
everything else stays the same.

```{r}
#| label: sol8-data
penguins_large <- tibble(
  species     = c("Adelie", "Adelie", "Gentoo", "Gentoo"),
  island      = c("Biscoe", "Dream",  "Biscoe", "Biscoe"),
  body_mass_g = c(3750, 3800, 5200, 5400)
)
```

```{r}
#| label: sol8-new
# replace_when() takes: the column to modify, then condition ~ replacement pairs.
# Rows that don't match any condition keep their original value automatically.
penguins_large |>
  mutate(
    body_mass_g = replace_when(
      body_mass_g,
      body_mass_g > 5000 ~ 5000
    )
  )
```

**Why `replace_when()` is better:**

- No `.default = body_mass_g` boilerplate
- Clear that you're *modifying* an existing column, not building a new one
- Use multiple `condition ~ value` pairs to handle several replacements at once

---

# Quick Reference

| Exercise | Task | Old function | New function |
|----------|------|-------------|-------------|
| 1 | Drop rows matching all conditions | `filter(!(cond1 & cond2))` with NA guards | `filter_out(cond1, cond2)` |
| 2 | Keep rows matching any condition | `filter(a \| b)` | `filter(when_any(a, b))` |
| 3 | Complex grouped OR logic | `filter((a & b & c) \| (d & e))` | `filter(when_any(when_all(a, b, c), when_all(d, e)))` |
| 4 | Keep rows matching all conditions | `filter(a & b)` | `filter(when_all(a, b))` |
| 5 | Recode values (inline) | `mutate(case_when(x == "a" ~ "A", ...))` | `mutate(recode_values("a" ~ "A", ...))` |
| 6 | Recode values (lookup table) | Manual `case_when()` from vectors | `recode_values(from = ..., to = ...)` |
| 7 | Replace exact values in column | `case_when(x == "a" ~ "A", .default = x)` | `replace_values("a" ~ "A")` |
| 8 | Replace values by condition | `case_when(cond ~ val, .default = x)` | `replace_when(x, cond ~ val)` |
